# Exploratory data analysis on US predential elections

### By Poonam

================================================================================


This is an exploration of 2016 US presidential election donations in the state 
of California. For this data analysis, I am exoloring the 2016 presidential 
campaign finance data from Federal Election Commission. The dataset contains 
financial contribution transaction.


Through my analysis, I will attempt to answer the following questions:

- Which candidate received the most money?
- Which political party received the most contributions?
- What is the spread of occupation of those donors?
- I would also see how the money flow happened from begin till end.
- Did Hillary Clinton receive more money than Donald Trump?


```{r Library Packages, message = FALSE, warning = FALSE}
# Load all of the packages that will be used for analysis
library(readr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(lubridate)
library(gridExtra)
library(plotly)
library(ggmap)
library(maps)
library(tidyverse)
```


```{r Load_the_Data, message = FALSE, warning = FALSE, include = FALSE, echo = FALSE}
# Load the Data
# include=FALSE - No code printed, no result printed
# echo=FALSE - No code printed, only result printed
# results = "hide" - Code Printed, No result printed
getwd()
list.files()
setwd("/Users/poonam/Documents/Poonam/DAND/DE-Project3/")
DF = read_csv('P00000001-CA_test.csv') 
```


Summarise the dataset, and check column names.


```{r Dataset Summary Column Names, message = FALSE, warning = FALSE, echo = FALSE}
# str(DF)  # data of first 10 rows
# dim(DF)  #number of observations and column count
summary(DF) #lenght of each column, min,max, median of numeric columns
colnames(DF)
```


# Univariate Plot Section

From the dataset summary we found that this dataset contains 1304346 
observations and 18 variables. Let's plot contribution graphs against different 
variables.

Let's do some basic null checks, before we proceed.


```{r null_row_check, message = FALSE, warning = FALSE, echo = FALSE, results="hide"}
null_rows = which(is.na(DF$contb_receipt_amt), arr.ind=TRUE)
null_row_count = length(null_rows) 
print(null_row_count) 
```


There are no null values under contribution amount column. Let's plot a simple 
histogram.


```{r Simple Histogram, message = FALSE, warning = FALSE, echo=FALSE}
hist(DF$contb_receipt_amt, 
     main="Histogram for contributions", 
     breaks=(25), 
     xlab="Contribution Amount in Dollars",
     ylab="Number of Contributions",
     border="Green", 
     col="Blue",
     xlim=c(-12000,12000),
     ylim =c(0, 150000))
```


Plot another detailed contribution histogram.


```{r  Bar chart for contributions, message = FALSE, warning = FALSE, echo=FALSE}
options(scipen = 999)
plot_cand_contb= ggplot(aes(x=contb_receipt_amt), data=(DF))+
  geom_histogram(fill='black') + 
  scale_x_continuous(limits=c(0, 3000), breaks = c(0, 250, 500, 1000, 1500, 
  2000, 2500, 3000))+ 
  scale_y_continuous(limits = c(0, 225000),
                     breaks = c(100000, 200000,300000)) + 
  labs(title="Contribution Distribution") +
  xlab("Contribution Amount in Dollars") +
  ylab("Number of Contributions")

plot_cand_contb
```


This graph also shows that large number of people have contributed below $500. 
Also there is a significant number of people those who have contributed between 
2500 and 3000. 

We can see from the above histogram, that there are negative values also in the 
contribution amount. Let's see if they have any other relevant details. 
Check their receipt description field.


```{r Negative amount desc, message = FALSE, warning = FALSE, echo=FALSE }
DF_neg_amt_desc <- subset(DF, select=c("cand_nm", "contbr_nm", "contbr_city",       "contbr_occupation", "contb_receipt_amt", "contb_receipt_dt", "receipt_desc", 
  "memo_text", "election_tp"), DF$contb_receipt_amt <0)
DF_neg_amt_desc

num_refund_rows <- which(DF_neg_amt_desc$receipt_desc == "Refund")
refund_row_count = length(num_refund_rows) 
print(refund_row_count)

unique_receipt_desc <- length(unique(DF_neg_amt_desc$receipt_desc, 
  incomparables = FALSE, MARGIN = 1, fromLast = FALSE))
print(unique_receipt_desc)
```


There are lot of records (10,412 out of 16,313), that have receipt description 
as "Refund". It could be that the contributor changed his mind, and asked for a 
refund at a later stage. Other 2 categories of description are "Redesignation" 
and "Reattribution". Let's see how significant is the sum of negative amount
in total.


Lets do some calculation to see the total sum of positive and negative values.

Check number of rows containing positive amount, and number of rows containing 
negative amount.


```{r Positive and Negative Contribution, message = FALSE, warning = FALSE, echo=FALSE, results= "hide"}

pos_row_count = length(which(DF$contb_receipt_amt>=0, arr.ind = TRUE)) 
print(pos_row_count)

neg_row_count = length(which(DF$contb_receipt_amt < 0, arr.ind = TRUE)) 
print(neg_row_count)
```

So out of 1304346 observations, 16313 are with negative amount values.

Lets see how the contribution is done over a period of time.
We will plot a time series line chart for each party to see trend of 
contribution received towards begin or end of the election time.

Let's do a basic null check on date before we proceed


```{r Null row check for date, message = FALSE, warning = FALSE, echo=FALSE}
null_rows = which(is.na(DF$contb_receipt_dt), arr.ind=TRUE)
null_row_count = length(null_rows)
print(null_row_count)
```


There are no rows without date field.
Check how many unique date records are present.


```{r Unique dates, message = FALSE, warning = FALSE}
count_of_contb_dt <- length(unique(DF$contb_receipt_dt, incomparables = FALSE, 
  MARGIN = 1, fromLast = FALSE))
print(count_of_contb_dt)
```


So the contribution distribution is spread across more than 2 years. 
Let's group the contribution by month.


```{r Add Political Party Column, message = FALSE, warning = FALSE, echo=FALSE}
DF["Political_Party"] <- NA
DF$Political_Party <- ifelse(DF$cand_nm == "Clinton, Hillary Rodham", 
                                            "Democratic_Party",
                       ifelse(DF$cand_nm == "Sanders, Bernard", 
                                             "Democratic_Party",
                         ifelse(DF$cand_nm == "Lessig, Lawrence", 
                                             "Democratic_Party",
                           ifelse(DF$cand_nm == "Malley, Martin Joseph", 
                                              "Democratic_Party",
                             ifelse(DF$cand_nm == "Webb, James Henry Jr", 
                                              "Democratic_Party",
                               ifelse(DF$cand_nm == "Johnson, Gary", 
                                              "Libretarian_Party",
                                 ifelse(DF$cand_nm == "Stein, Jill", 
                                              "Green_Party_of_USA", 
                                              "Republic_Party")))))))
```



```{r Date coversion, message = FALSE, warning = FALSE, echo = FALSE, results= "hide"}
DF$Month_Yr <- format(as.Date(DF$contb_receipt_dt, "%d-%b-%y"), "%b-%Y")
DF$yyyymm <- format(as.Date(DF$contb_receipt_dt, "%d-%b-%y"), "%Y%m")
DF

DF_sum_by_month <- subset(DF, select=c("Political_Party", "contb_receipt_amt", "Month_Yr","yyyymm"))
DF_sum_by_month

DF_sum_by_month <- arrange(DF_sum_by_month, yyyymm)  
DF_sum_by_month

DF_sum_by_month <- DF %>% 
  group_by(yyyymm) %>% 
  summarise(sum_contb_by_month = sum(contb_receipt_amt, na.rm=TRUE),
  mean_occu = mean(contb_receipt_amt),
  n = n())

DF_sum_by_month$Month_Yr <- as.Date(paste0(as.character(DF_sum_by_month$yyyymm),
  '01'), format='%Y%m%d')

DF_sum_by_month$Month_Yr <- format(as.Date(DF_sum_by_month$Month_Yr, "%Y%m%d"), 
  "%b-%Y")

DF_sum_by_month_head <- head(arrange(DF_sum_by_month, desc(sum_contb_by_month)), n=20)
DF_sum_by_month_head
```


The top 10 dates of contribution show Dec 2015 being the peak month of 
contribution, followed by Sep 2015.

Let's visualize through a graph. 


```{r Plot by Date, message = FALSE, warning = FALSE, echo= FALSE }

plot_contb_by_month <- ggplot(data = DF_sum_by_month_head, 
  aes(x = DF_sum_by_month_head$yyyymm, 
  y = (DF_sum_by_month_head$sum_contb_by_month/1000))) +
  geom_bar(stat="identity")+
  geom_line() +
  theme(axis.text.x=element_text(angle=45,hjust=1,vjust=1))+
  xlab('Date in yyyymm format') +
  ylab('Contribution Amount in Dollars (Thousands)') +
  ggtitle('Contribution by Month') +
  scale_y_continuous(limits = c(0, 20000))

plot_contb_by_month
```


# Univariate Analysis


In the Univariate section we explored the "Contribution Amount" variable. We saw
that contribution was mostly between the range of 0 to 500 Dollar. Also 
there was a slight peak at 1000 and 2700 dollar. There were mostly positive 
contribution values, but there were some observations with negative values too. 
The negative values had description as Refund, Reattribution and Redesignation. 
For the purpose of current exploration the total amount is calculated as sum of 
positive amount minus the sum of negative amount.

There are 1304346 contributions and 18 variables. The variables that interest 
to me and I have used are:

cand_nm: Candidate Name
contb_receipt_amt: Contribution Amount
contbr_occupation: Contributor Occupation
contbr_city: Contributor City
contb_receipt_dt: Contribution date
election_tp: Type of election (Primary, General)

Othere observations:

Most people contribute small amount of money. The median contribution amount is 
$27, mean contribution amount is $116. The amount of contribution is highest in 
Aug-Sep 2016, that is just before the election.

From the above graph we see that contribution was maximum between Aug 2016 till 
Oct 2016, just before the election time frame. Also we can note that the 
contribution started to pick from Apr 2015.


# Bivariate Plot Section


Next we will see how this total "contribution amount" is distributed w.r.t other 
factors like  candidates, political parties, contributor's occupation and 
contributor's city.


Get the Unique candidate names to see how many candidates stood up for election.
Summarise the total contribution amount for each candidate.


```{r Number of unique candidates, message = FALSE, warning = FALSE}
count_of_nm <- length(unique(DF$cand_nm, incomparables = FALSE, MARGIN = 1, 
  fromLast = FALSE))
print(count_of_nm)
print(unique(DF$cand_nm)) 
```


We see that there are 25 unique candidtes.
Plot a bar chart to see how much contribution amount is received per candidate.


```{r Candidate wise contribution, message = FALSE, warning = FALSE}
DF_cand_dist <- DF %>% 
  group_by(cand_nm) %>% 
  summarise(candidate_amt = sum(contb_receipt_amt, na.rm=TRUE),
            n= n())
DF_cand_dist

DF_cand_dist <- head(arrange(DF_cand_dist, desc(candidate_amt)), n= 10)
```


Let's plot the graph.


```{r Plot by Candidate, message = FALSE, warning = FALSE, echo = FALSE}
plot_cand_dist <- ggplot(DF_cand_dist, aes(x=DF_cand_dist$cand_nm, 
  y = candidate_amt/1000)) +
  geom_col(inherit.aes = TRUE)+
  theme(axis.text.x=element_text(angle=45,hjust=1,vjust=1))+
  ylab("Contribution Amount in Dollars (Thousands)") +
  xlab("Candidate Names")+
  ggtitle('Contribution Received by Candidate')
plot_cand_dist
```



Check political party wise contribution.


Draw pie chart to see distribution of contribution amount received by each 
political party.

First sum the contribution party wise.


```{r Political Party Contribution, message = FALSE, warning = FALSE}
DF_party_dist <- DF %>% 
  group_by(Political_Party) %>% 
  summarise(party_amt = sum(contb_receipt_amt, na.rm=TRUE))

head(DF_party_dist)
```


```{r Pie chart, message = FALSE, warning = FALSE, echo = FALSE, results="hide"}
DF_dem_new = subset(DF, Political_Party == "Democratic_Party")
dem_sum = sum(DF_dem_new$contb_receipt_amt)
print(dem_sum)

DF_lib_new = subset(DF, Political_Party == "Libretarian_Party")
lib_sum = sum(DF_lib_new$contb_receipt_amt)
print(lib_sum)

DF_gpusa_new = subset(DF, Political_Party == "Green_Party_of_USA")
gpusa_sum = sum(DF_gpusa_new$contb_receipt_amt)
print(gpusa_sum)

DF_rep_new = subset(DF, Political_Party == "Republic_Party")
rep_sum = sum(DF_rep_new$contb_receipt_amt)
print(rep_sum)

slices = c(sum(DF_dem_new$contb_receipt_amt), 
           sum(DF_lib_new$contb_receipt_amt),
           sum(DF_rep_new$contb_receipt_amt),
           sum(DF_gpusa_new$contb_receipt_amt)
           )
colors <- c("blue","white","red","green")
slices_labels <- round(slices/sum(slices) * 100, 1)
slices_labels <- paste(slices_labels, "%", sep="")

pie(slices, main="Political Party wise Contribution", col=colors, 
  labels=slices_labels, cex=0.8)
legend("topright", c("Democratic Party","Libretarian Party","Republic Party", 
  "Green Party of USA"), cex=0.8, fill=colors)
```


Lets see how many election types are there.


```{r Contribution per election type, message = FALSE, warning = FALSE }
count_of_election_tp <- length(unique(DF$election_tp, incomparables= FALSE,
  MARGIN = 1, fromLast = FALSE))
print(count_of_election_tp)
print(unique(DF$election_tp))
```


We see that there are 5 unique election types. Let's see contribution per 
election type.


```{r Election type wise contribution, message = FALSE, warning = FALSE}
DF_election_tp <- DF %>% 
  group_by(election_tp) %>% 
  summarise(sum_election_tp = sum(contb_receipt_amt, na.rm=TRUE),
                              mean_election_tp = mean(contb_receipt_amt), 
                              n = n())
DF_election_tp <- head(arrange(DF_election_tp,desc(sum_election_tp)))
head(DF_election_tp)
```


```{r Election Type Plot, message = FALSE, warning = FALSE, echo=FALSE}
plot_election_tp <- ggplot(DF_election_tp, aes(x=DF_election_tp$election_tp, y =                  sum_election_tp/1000)) +
  geom_bar(stat="identity")+
  geom_col(inherit.aes = TRUE)+
  theme(axis.text.x=element_text(angle=45,hjust=1,vjust=1))+
  ylab("Contribution Amount in Dollars (Thousands)") +
  xlab("Type of Election")+
  ggtitle('Contribution by Election Type')

plot_election_tp
```


Most of the contribution is for election type P2016. This could be the Primary 
election donation. The next type of election that has received most 
contributions is G2016, this could be the general elections. Source of 
information wikipedia website.


```{r Candidate and election type,  message = FALSE, warning = FALSE, echo=FALSE}
DF_election_tp_p2016 <- subset(DF, select=c("cand_nm", "contbr_nm", "contbr_city",       "contbr_occupation", "contb_receipt_amt", "contb_receipt_dt", "receipt_desc", 
  "memo_text", "election_tp"), DF$election_tp == "P2016")
DF_election_tp_p2016

DF_election_tp_p2016_grp<- DF_election_tp_p2016 %>% 
  group_by(cand_nm) %>% 
  summarise(sum_election_tp_p2016 = sum(contb_receipt_amt, na.rm=TRUE),
                              mean_election_tp = mean(contb_receipt_amt), 
                              n = n())
DF_election_tp_p2016_grp <- head(arrange(DF_election_tp_p2016_grp
                                         ,desc(sum_election_tp_p2016)))
DF_election_tp_p2016_grp
summary(DF_election_tp_p2016_grp)
```


```{r P2016 contribution distribution, message = FALSE, warning = FALSE, echo=FALSE}
plot_election_tp_p2016_grp <- ggplot(DF_election_tp_p2016_grp, 
  aes(x=DF_election_tp_p2016_grp$cand_nm, y = sum_election_tp_p2016/1000)) +
  geom_bar(stat="identity")+
  geom_col(inherit.aes = TRUE)+
  theme(axis.text.x=element_text(angle=45,hjust=1,vjust=1))+
  ylab("Pri Election Amount in Dollars (Thousands)") +
  xlab("Candidate Name")+
  ggtitle('Contribution Received by Candidates for P2016')

plot_election_tp_p2016_grp
```


Above is a graph for primary election contribution amount by candidates. Hillary
got most contribution during the primary elections as compared to other 
candidates, followed by Bernard Sanders. The mean contribution amount is $318.1 
for primary elections, this is more than the mean of total contribution amount 
that is $116.


```{r Number of unique occupations, message = FALSE, warning = FALSE}
count_of_occupation <- length(unique(DF$contbr_occupation, incomparables= FALSE,
  MARGIN = 1, fromLast = FALSE))
print(count_of_occupation)
```


We see that there are 28616 unique occupations of the contributors. People from 
so many occupations participated in contributing to the election. It would be 
difficult to see contribution spread against all of the occupation. 
Lets pick top 10 occupation categories.


```{r Occupation wise contribution, message = FALSE, warning = FALSE}
DF_occu_dist <- DF %>% 
  group_by(contbr_occupation) %>% 
  summarise(sum_occup = sum(contb_receipt_amt, na.rm=TRUE),
                              mean_occu = mean(contb_receipt_amt), 
                              n = n())
DF_occu_dist <- head(arrange(DF_occu_dist,desc(sum_occup)), n = 10)
DF_occu_dist
summary(DF_occu_dist)
```


From the summary we see that the occupation categories "ATTORNEY", "HOMEMAKER", 
"CEO", "PRESIDENT", "LAWYER" have the higher mean of contribution than compared 
to the mean of total contributing amount from all categories together.

Category "PRESIDENT" has the maximum mean of contributing amount.
Category "NOT EMPLOYED" has the minimum mean of contributing amount.


```{r Occupation Plot, message = FALSE, warning = FALSE, echo=FALSE}
plot_occu_dist <- ggplot(DF_occu_dist, aes(x=DF_occu_dist$contbr_occupation, y =                  sum_occup/1000)) +
  geom_bar(stat="identity")+
  geom_col(inherit.aes = TRUE)+
  theme(axis.text.x=element_text(angle=45,hjust=1,vjust=1))+
  ylab("Contribution Amount in Dollars (Thousands)") +
  xlab("Contributor's Occupation")+
  ggtitle('Contribution by Occupation')
  
plot_occu_dist
```


This is an amazing graph. It shows retired category of people contributing most 
to the election funds.

Check city wise contribution, and then find out top 10 contributing cities.
Below is graph for top 10 contributing cities.


```{r Count of Unique cities, message = FALSE, warning = FALSE}
count_of_city <- length(unique(DF$contbr_city, incomparables = FALSE, 
  MARGIN = 1, fromLast = FALSE))
print(count_of_city)
```


```{r Top contributing cities, message = FALSE, warning = FALSE}
DF_contbr_city <- DF %>% 
  group_by(contbr_city) %>% 
  summarise(sum_city = sum(contb_receipt_amt, na.rm=TRUE),
                              mean_city = mean(contb_receipt_amt), 
                              n = n())
DF_contbr_city <- head(arrange(DF_contbr_city,desc(sum_city)), n = 10)
DF_contbr_city
```


```{r Top contributor cities graph, message= FALSE, warning= FALSE, echo= FALSE}
plot_contbr_city <- ggplot(DF_contbr_city, aes(x=DF_contbr_city$contbr_city, 
  y = sum_city/1000)) +
  geom_bar(stat="identity")+
  geom_col(inherit.aes = TRUE)+
  theme(axis.text.x=element_text(angle=45,hjust=1,vjust=1))+
  ylab("Contribution Amount in Dollars (Thousands)") +
  xlab("Contributor's City")+
  ggtitle('Contribution by City')

plot_contbr_city
```


Los Angeles is the most contributing city out of all, followed closely by 
San Francisco. We see that the "amount of contribution" from these 
cities is more, but is the "count of contributions" also more from these cities.
Let's see number of contributions per candidate, occupation and city.

Group contribution amount per candidate, per occupation, per city.


```{r Count of Contributions, message= FALSE, warning= FALSE}
options(scipen = 999)
DF_cand_occu_city_grp <- DF %>% 
   group_by(.dots=c("cand_nm","contbr_occupation","contbr_city")) %>% 
   summarise(sum_cand_occu_city=sum(contb_receipt_amt),
             n = n())

DF_cand_occu_city_grp <- (arrange(DF_cand_occu_city_grp,desc(sum_cand_occu_city)))
DF_cand_occu_city_grp
```


```{r Graph for grouped data,  message= FALSE, warning= FALSE, echo = FALSE}
top_candidate <- c("Clinton, Hillary Rodham", 
                   "Sanders, Bernard", 
                   "Trump, Donald J.", 
                   "Cruz, Rafael Edward 'Ted'", 
                   "Rubio, Marco",
                   "Bush, Jeb",
                   "Carson, Benjamin S.",
                   "Kasich, John R.",
                   "Fiorina, Carly",
                   "Paul, Rand")
top_candidate

top_candidate1 <- c("Clinton, Hillary Rodham", 
                   "Sanders, Bernard", 
                   "Trump, Donald J.", 
                   "Cruz, Rafael Edward 'Ted'", 
                   "Rubio, Marco")


top_occu <- c("RETIRED", 
              "ATTORNEY", 
              "NOT EMPLOYED", 
              "INFORMATION REQUESTED",
              "HOMEMAKER",
              "CEO",
              "PHYSICIAN",
              "CONSULTANT",
              "PRESIDENT",
              "LAWYER")
top_occu

top_city <- c("LOS ANGELES", 
              "SAN FRANCISCO", 
              "SAN DIEGO", 
              "PALO ALTO", 
              "OAKLAND",
              "BEVERLY HILLS",
              "BERKELEY",
              "SANTA MONICA",
              "SAN JOSE",
              "SACRAMENTO")
top_city

plot_count_cand <- ggplot(DF_cand_occu_city_grp, 
  aes(x=as.factor(DF_cand_occu_city_grp$cand_nm), y = n)) +
  geom_bar(stat="identity")+
  geom_col(inherit.aes = TRUE)+
  theme(axis.text.x=element_text(angle=45,hjust=1,vjust=1))+
  scale_x_discrete(limits = c(top_candidate), breaks = c(top_candidate))+
  ylab("Contribution count per Candidate") +
  xlab("Candidate Name")+
  ggtitle('Number of Contributions Received by Candidate')

plot_count_cand


plot_count_occu <- ggplot(DF_cand_occu_city_grp, 
  aes(x=as.factor(DF_cand_occu_city_grp$contbr_occupation), y = n)) +
  geom_bar(stat="identity")+
  geom_col(inherit.aes = TRUE)+
  theme(axis.text.x=element_text(angle=45,hjust=1,vjust=1))+
  scale_x_discrete(limits = c(top_occu), breaks = c(top_occu))+
  ylab("Count of Contributions per Occupation") +
  xlab("Contributor's Occupation")+
  ggtitle('Number of Contributions Received by Occupation')

plot_count_occu


plot_count_city <- ggplot(DF_cand_occu_city_grp, 
  aes(x=as.factor(DF_cand_occu_city_grp$contbr_city), y = n)) +
  geom_bar(stat="identity")+
  geom_col(inherit.aes = TRUE)+
  theme(axis.text.x=element_text(angle=45,hjust=1,vjust=1))+
  scale_x_discrete(limits = c(top_city), breaks = c(top_city))+
  ylab("Contribution count per City") +
  xlab("Contributor's City")+
  ggtitle('Number of Contributions Received by City')
  
plot_count_city
```


In the above graph the occupation categories are placed as per the amount of 
donations they have made.
The most significant observation is between "Attorney" and  "Not Employed". 
We see, that though more number of "Not Employed" people have donated but there 
amount of contribution was less than the amount of contributions done by 
category "Attorney".

On the contribution city graph also the cities are arranged in the descending
order of their amount of contributions done.
The most significant point is where amount of contributions done by city 
"Palo Alto" is more than "Oakland", but here in the graph the number of 
contributions done by city of "Palo Alto" is much lesser than "Oakland".

This may suggest that more financially sound people stay in "Palo Alto".


```{r Correlation, message= FALSE, warning= FALSE, results="hide"}

with(DF, cor(contb_receipt_amt, rank(cand_nm)))
with(DF, cor(contb_receipt_amt, rank(contbr_occupation)))
with(DF, cor(contb_receipt_amt, rank(contbr_city)))

with(DF, cor(rank(cand_nm), rank(contbr_occupation)))
with(DF, cor(rank(cand_nm), rank(contbr_city)))
with(DF, cor(rank(contbr_occupation), rank(contbr_city)))

with(DF_cand_occu_city_grp, cor(sum_cand_occu_city, n))
```


There doesn't seem to be much correlation between candidate name and 
contributor city, or between contributor city and occupation.
One thing that shows a strong uphill correlation is sum of contrubition amount 
and the number of contributions.
That is more the  "number of contributions" per candidate, per occupation, 
per city , more is the value of contribution.


```{r Retired per city LA, message=FALSE, warning=FALSE, echo = FALSE, results="hide"}
# What occupation does people have mostly as per most contributing city
retired_count_total = length(which(DF$contbr_occupation == "RETIRED")) 
print(retired_count_total)

total = 1304346
percentage_of_retired_total = (retired_count_total/total)*100
print(percentage_of_retired_total)

retired_count_LA = length(which(DF$contbr_occupation == "RETIRED" & 
                          DF$contbr_city == "LOS ANGELES")) 
print(retired_count_LA)
total_LA = length(which(DF$contbr_city == "LOS ANGELES")) 
print(total_LA)
percentage_retired_LA = (retired_count_LA/total_LA)*100
print(percentage_retired_LA)

```

Percentage of "Retired" category of the total contributors is 20%. That is 
almost 1/5th of the total contributors are Retired Category.

Let's check the retired percentage for Los Angeles and San Francisco.

We calculated that, roughly 10% of all the contributors from Los Angeles are of "Retired" 
category. This is half of total percentage.


```{r Retired per city SF, message=FALSE, warning=FALSE, echo = FALSE, results="hide"}
retired_count_SF = length(which(DF$contbr_occupation == "RETIRED" & 
                          DF$contbr_city == "SAN FRANCISCO")) 
print(retired_count_SF)
total_SF = length(which(DF$contbr_city == "SAN FRANCISCO")) 
print(total_SF)
percentage_retired_SF = (retired_count_SF/total_SF)*100
print(percentage_retired_SF)
```


Similar to Los Angeles, from San Francisco also 10% of all the contributors 
are of "Retired" category. This data shows the percentage of "Retired" category
contributors is quite significant in other cities also, and not in LA and 
SF only.

# Bivariate Analyasis

For the bivariate analysis I saw how the total contribution amount is 
distributed w.r.t following factors.

- Candidates
- Political Party
- Contributor's Occupation
- Contributor's City
- Election Type

There is no mention of political party in the dataset, this was added to the
dataset.
I added the column political_party and filled the column with corresponding 
party name for each candidate's political party. Used website 
http://www.politifact.com/ to get info on 25 unique candidates.
I found that 25 candidates belonged to 4 different political parties, namely

- Democratic Party
- Republic Party
- Libretarian Party
- Green Party of USA

Hillary Clinton received the most contribution. 
That also reflected in the political party she represented.
Democratic party received the most contribution almost 75% of the total 
contributed amount. 
Retired people contributed the most 
Most contributing cities were Los Angeles and San Francisco.
Most of the contributions were made for primary election.


# Multivariate Plot Section

Let's plot a map diagram, to see location of the most contributing cities,
on the California map.


```{r city_map, message=FALSE, warning=FALSE, echo = FALSE}
DF_contbr_city_20 <- DF %>% 
  group_by(contbr_city) %>% 
  summarise(sum_city = sum(contb_receipt_amt, na.rm=TRUE),
                              mean_amt_per_city = mean(contb_receipt_amt), 
                              n = n())
DF_contbr_city_20 <- head(arrange(DF_contbr_city_20,desc(sum_city)), n = 20)

DF_contbr_city_20 <- cbind(geocode(as.character(DF_contbr_city_20$contbr_city)), 
                            DF_contbr_city_20)
DF_contbr_city_20

city_contb_map <- ggmap(get_map(location = 'California', zoom = 7)) +
  geom_point(data=DF_contbr_city_20, 
             aes(x=lon, y=lat, size=n), 
             color="orange",
             na.rm = TRUE)+
             ylab("Latitude") +
             xlab("Longitude")+
            ggtitle('Contribution by top Cities')
city_contb_map
```


In the above graph, size of the orange dot specifies the count of contribution.

Let's plot some heat maps, to see the multivariate effect on the contributions.
We will see contributions for candidates w..r.t cities and occupations.  So far 
we know most contributing occupation, but we will see most contributing 
occupation per city through some heat maps.



```{r heat_map1, message=FALSE, warning=FALSE, echo = FALSE}
# HeatMap 1 - cand-city-contribution
options(scipen = 999)
DF_cand_city_grp <- DF %>% 
   group_by(.dots=c("cand_nm","contbr_city")) %>% 
   summarise(sum_cand_city=sum(contb_receipt_amt),
             n = n())

DF_cand_city_grp <- (arrange(DF_cand_city_grp,desc(sum_cand_city)))
DF_cand_city_grp

group = c(DF_cand_city_grp$cand_nm)
pos = c(DF_cand_city_grp$contbr_city)
color = c(DF_cand_city_grp$n)
df = data.frame(group, pos, color)
ggplot(data = df, aes(x = group, y = pos)) + 
          geom_tile(aes(fill = color))+
          scale_x_discrete(limits = c(top_candidate), breaks = c(top_candidate))+
          scale_y_discrete(limits = c(top_city), breaks = c(top_city))+
          theme(axis.text.x=element_text(angle=45,hjust=1,vjust=1))+
          theme(axis.text.y=element_text(hjust=1,vjust=1))+
          xlab("Candidate Name")+
          ylab("Contributor's city") +
          ggtitle('Contribution Received by Candidates per city')+
          scale_fill_gradient(low="pink", high="blue")
```



```{r heat_map2, message=FALSE, warning=FALSE, echo = FALSE}
# HeatMap 2 - cand-occu-contribution
DF_cand_occu_grp <- DF %>% 
   group_by(.dots=c("cand_nm","contbr_occupation")) %>% 
   summarise(sum_cand_occu=sum(contb_receipt_amt),
             n = n())
DF_cand_occu_grp <- (arrange(DF_cand_occu_grp,desc(sum_cand_occu)))
DF_cand_occu_grp

group = c(DF_cand_occu_grp$cand_nm)
pos = c(DF_cand_occu_grp$contbr_occupation)
color = c(DF_cand_occu_grp$n)
df = data.frame(group, pos, color)
ggplot(data = df, aes(x = group, y = pos)) + 
          geom_tile(aes(fill = color))+
          scale_x_discrete(limits = c(top_candidate), breaks = c(top_candidate))+
          scale_y_discrete(limits = c(top_occu), breaks = c(top_occu))+
          theme(axis.text.x=element_text(angle=45,hjust=1,vjust=1))+
          theme(axis.text.y=element_text(hjust=1,vjust=1))+
          ylab("Contributor's Occupation") +
          xlab("Candidate Name")+
          ggtitle('Contribution Received by Candidate per Occupation')+
          scale_fill_gradient(low="pink", high="blue")
```



```{r heat_map3, message=FALSE, warning=FALSE, echo = FALSE}
# HeatMap 3 - occu-city-contribution
DF_occu_city_grp <- DF %>% 
   group_by(.dots=c("contbr_occupation","contbr_city")) %>% 
   summarise(sum_cand_occu=sum(contb_receipt_amt),
             n = n())
DF_occu_city_grp <- (arrange(DF_occu_city_grp,desc(sum_cand_occu)))
DF_occu_city_grp

group = c(DF_occu_city_grp$contbr_occupation)
pos = c(DF_occu_city_grp$contbr_city)
color = c(DF_occu_city_grp$n)
df = data.frame(group, pos, color)
ggplot(data = df, aes(x = group, y = pos)) + 
          geom_tile(aes(fill = color))+
          scale_x_discrete(limits = c(top_occu), breaks = c(top_occu))+
          scale_y_discrete(limits = c(top_city), breaks = c(top_city))+
          theme(axis.text.x=element_text(angle=45,hjust=1,vjust=1))+
          theme(axis.text.y=element_text(hjust=1,vjust=1))+
          xlab("Contributor's Occupation")+
          ylab("Contributor's City") +
          ggtitle('Contribution Received by City per Occupation')+
          scale_fill_gradient(low="pink", high="blue")
```


The above heatmaps show, Hillary having good number of contributions from
Los Angeles and San Francisco. Only top two candidates have more than 2000,
contributions from most of the cities, rest of the candidates have less than 
2000 contributions, across all cities.

On the occupation wise map also Hillary has got contributions across all
occupations.

"Not Employed" category of people have contributed most to Bernard Sanders.


```{r time_series, message=FALSE, warning=FALSE, echo = FALSE}

DF_cand_month_grp <- DF %>% 
  group_by(cand_nm, yyyymm) %>% 
  summarise(sum_contb_cand_month = sum(contb_receipt_amt, na.rm=TRUE),
  mean_occu = mean(contb_receipt_amt),
  n = n())

plot_cand_dt_amt_time_series <- ggplot(data=DF_cand_month_grp[!is.na(DF_cand_month_grp$sum_contb_cand_month),],
    aes(x = yyyymm, y = sum_contb_cand_month/1000, color = cand_nm)) +
    geom_path()+
    geom_line()+
    geom_point()+
    xlab('Date in yyyymm') +
    ylab('Contribution Amount in Dollars (Thousands)') +
    theme(axis.text.x=element_text(angle=45,hjust=1,vjust=1))+
    ggtitle('Time Series of Contribution Amount by Candidate')

plot_cand_dt_amt_time_series

```



# Multivariate Analysis

We know that Hillary Clinton raised the most money and had the most supporters 
in California. But is this always true throughout the campaign process? 
When we look at above 2 graphs, we can notice few things.

Hillary Clinton had most number of contributions throughout.
Number of contributions for Bernard Sanders rose quite consistently.
Number of contributions for Donald Trump fell towards the end of campaign.
Towards the end only Bernard Sanders was in some competition to Hillary Clinton
in terms of number of contributions.


# Final Plots and Summary


### Plot One


```{r  Final Plot 1 contribution, message = FALSE, warning = FALSE, echo=FALSE}
options(scipen=999)
plot_cand_contb= ggplot(aes(x=contb_receipt_amt), data=(DF))+
  geom_histogram(fill='black') + 
  scale_x_continuous(limits=c(0, 3000), breaks = c(0, 250, 500, 1000, 1500, 
  2000, 2500, 3000))+ 
  scale_y_continuous(limits = c(0, 225000),
                     breaks = c(100000, 200000,300000)) + 
  labs(title="Contribution Distribution") +
  xlab("Contribution Amount in Dollars (Thousands)") +
  ylab("Number of Contributions")

plot_cand_contb
```


### Plot One description

This graph shows the count of contributions for each range of amount. 
Large number of people have made small donations between 0 to 250 dollars. 
Many contributions are done for the amount of 500, 1000 and 2700 Dollars.
From the summary we saw that mean contribution amount is 116 Dollars. 
This can be seen on the graph.


### Plot Two


```{r Final plot 2 candidate, message = FALSE, warning = FALSE, echo = FALSE}
plot_cand_dist <- ggplot(DF_cand_dist, aes(x=DF_cand_dist$cand_nm, 
  y = candidate_amt/1000)) +
  geom_col(inherit.aes = TRUE)+
  theme(axis.text.x=element_text(angle=45,hjust=1,vjust=1))+
  ylab("Contribution Amount in Dollars (Thousands)") +
  xlab("Candidate Names")+
  ggtitle('Contribution Received by Candidate')

plot_cand_dist
```


### Plot Two description

Hillary Clinton was the top candidate in terms of contibutions recieved. Her 
share of contribution was highest from the begin of primary election too. 
This graph answers my question that I thought of at the beginning of my 
exploration.


### Plot Three


```{r Final plot 3 Occupation, message = FALSE, warning = FALSE, echo=FALSE}
plot_occu_dist <- ggplot(DF_occu_dist, aes(x=DF_occu_dist$contbr_occupation, y =                  sum_occup/1000)) +
  geom_bar(stat="identity")+
  geom_col(inherit.aes = TRUE)+
  theme(axis.text.x=element_text(angle=45,hjust=1,vjust=1))+
  ylab("Contribution Amount in Dollars (Thousands)") +
  xlab("Contributor's Occupation")+
  ggtitle('Contribution by Occupation')
  
plot_occu_dist
```


### Plot Three description

This was one of the most interesting graph. Retired people contributed most to 
the 2016 election. We had also known that Los Angeles was the most contributing 
city. Does this mean that most of the retired category people stay at LA? This 
may not be a direct correlation, but something that can be explored. Also 
another correlation that can be thought of is that did Hillary receive most 
contribution from "Retired" category of people? 


# Reflection

This was a large dataset with more than a million and a quarter observations, 
which had details about the contributions made to political candidates during 
the 2016 US Presidential elections.

I was most interested to see which political party received the most funds. 
There was no political party column in the given dataset. I found the unique 
candidate names first and then searched for their parties, to finally see the 
pie chart for party wise funding. For the purpose of seeing the trend of 
contributions, on a time series, I had added two columns Month_Yr, and yyyymm.

The dataset was not a perfect clean data. I was getting parsing error while 
creating the dataframe. I found out that there was a extra comma character at 
the end of the last column, which was removed for successful parsing of csv file
. Also there were 7 columns where the zip code was non integer, example N4W2T. 
I found out that this zipcode belonged to Canada, and not California, USA. 
Such records were replaced by '000000000' value.

The most difficult decision for me was to handle the negative amount values. 
I did not want to ignore them initially. But after the entire exploration 
I realize that probably ignoring those values was a better choice. 
By the description it shows that, it is the contribution money to be refunded. 
It may not have reached the candidate/party at all. It was marked as negative, 
to be actually ignored.

For the future exploration I would like to see number of contributions and their 
respective contributors for large contribution amounts, above a certain average.

I could see total number of contributors per candidate, would like to see the 
data of number of contributors per candidate per city per occupation in one 
graph. I couldn't achieve more than 2 group by in one single graph.

During explortion I realized that most of the data in this dataset is 
categorical, except one continuous data that is "Contribution Amount". 
Rest most of the variables, that is candidate name, contributor occupation, 
contributor city, election type etc were discrete data points. 
So I have mostly plotted bar charts, and not scatter plots or line charts.

